Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-20 paper/conclusion.tex
adding period gap section and some rewriting

We found that old groups of K dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
% This indicates that either the \citep{angus2019} period-color relation is
% incorrect, or that mass-dependent dynamical heating is affecting this data set.
% We presented three pieces of evidence that suggest mass-dependent heating
% cannot be responsible for all of the increasing velocity dispersion.

@@

\racomment{Words about the significance of an inverted period-color relation.}

\racomment{How does the weakened van Saders braking law fit into all this?}

\racomment{Talk about the fact that the stars in each bin are not necessarily
the same age.}

\racomment{Dispersion could be dependent on the selection function.
We cut stars at high b, because we find that these do lead to a bias.
We find no evidence for increased velocity dispersion with decreasing teff in
either vb or vz.
We find that sigma clipping is necessary for vb but not vz and do not find
that the outliers are distributed towards higher galactic latitudes.
}

>>>>

We examined the gyrochronology relations using the velocity dispersions of
field stars with measured rotation periods.
We found that old groups of K dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
It appears that the period-\teff\ relation changes shape over time: rotation
period is negatively correlated with \teff\ at young ages, and positively
correlated at old ages.
The relation appears to start flattening out after $\sim$1 Gyr (see figure
\ref{fig:age_cut}), potentially around the same age, or just older than the
period gap which is located at a (gyro) age of around $\sim$1.1 Gyr.
This gap may represent a significant transitional epoch in the magnetic
behavior of stars.
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer and their rotation
periods can be measured at older ages.
% This indicates that either the \citep{angus2019} period-color relation is
% incorrect, or that mass-dependent dynamical heating is affecting this data set.
% We presented three pieces of evidence that suggest mass-dependent heating
% cannot be responsible for all of the increasing velocity dispersion.

@@

% \racomment{Words about the significance of an inverted period-color relation.}

% \racomment{How does the weakened van Saders braking law fit into all this?}

% \racomment{Talk about the fact that the stars in each bin are not necessarily
% the same age.}

% \racomment{Dispersion could be dependent on the selection function.
% We cut stars at high b, because we find that these do lead to a bias.
% We find no evidence for increased velocity dispersion with decreasing teff in
% either vb or vz.
% We find that sigma clipping is necessary for vb but not vz and do not find
% that the outliers are distributed towards higher galactic latitudes.
% }

### 2019-08-20 paper/method.tex
adding period gap section and some rewriting

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with negative parallaxes, parallax signal-to-noise ratios
less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
than 15\degrees\ from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$), we found that
including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
at low stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and keeping a large enough sample size to
provide meaningful results.
We also found that the anti-correlation between stellar mass and $b$ is
negligable for stars cooler than 5000 K.
For this reason, and because the rotational evolution of old, G stars does not
follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
dwarfs.

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different age ranges \citep[where age was
calculated using the][gyrochronology relation]{angus2019}, and calculated the
velocity dispersion: the standard deviation of velocities, \sigmavb, as a
function of effective temperature for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of {\it effective
temperature}.
We chose to use effective temperatures and not colors in this analysis as it
is the linear quantity and therefore easier to divide into bins of roughly
equal numbers of stars.

>>>>

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with negative parallaxes, parallax signal-to-noise ratios
less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
than 15\degrees\ from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$), we found that
including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
at low stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and keeping a large enough sample size to
provide meaningful results as we found that reducing this to 10\degrees did
not significantly affect our results.
We also found that the anti-correlation between stellar mass and $b$ is
negligable for stars cooler than 5000 K.
For this reason, and because the rotational evolution of old, G stars does not
follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
dwarfs.

### 2019-08-20 paper/results.tex
adding period gap section and some rewriting

@@

@@

@@

% If we assume the rise in velocity dispersion at cooler effective temperatures
% is caused by an inaccurate period-color relation rather than mass-dependent
% dynamical heating, we can generate a graphical representation of what the
% relations {\it should} look like.
\begin{figure}
  \caption{
Rotation period vs effective temperature for stars in the \mct\ sample,
    colored by the velocity dispersions of stars in restricted period and
    temperature ranges.
}
  \centering
    \includegraphics[width=1\textwidth]{dispersion_period_teff}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows the stars in our sample of
rotating dwarfs on a
plot of period vs effective temperature, coloured by their (\vb) velocity
dispersion.
Here, velocity dispersion was calculated for groups of stars in
uniformly-spaced bins in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating doesn't affect this sample, and \vb\
at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, it appears as though rotation period {\it decreases}
with decreasing effective temperature at old ages.  This is a paradigm shift
for gyrochronology because stellar spin-down rate is thought to be directly
tied to magnetic field strength, and the deeper convection zones of cooler
stars generate stronger magnetic fields which {\it should} lead to more
efficient angular momentum loss.

Something like this has been observed in the 1.1 Gyr NGC 6811 cluster, where
the rotation periods of mid-K dwarfs are faster than expected; their
rotational evolution appears to have stalled.
NGC 6811 straddles the rotation period gap and may be the `missing link' of
gyrochronology that connects the two periods of stellar spin-down: the early
stage where the period-color relation for K dwarfs has a positive slope and
the late stage where it has a negative slope.
In this case the period gap may delineate the transition between these two
regimes and is the point at which stellar magnetic dynamos likely undergo a
dramatic structural shift.

Figure \ref{fig:dispersion_period_teff} also indicates that the oldest stars
with measured rotation periods are cool, which may be because cooler stars
remain active for longer, and their rotation periods can therefore be measured
at older ages.

>>>>

\subsection{Velocity dispersion of coeval groups}

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different age ranges \citep[where age was
calculated using the][gyrochronology relation]{angus2019}, and calculated the
velocity dispersion: the standard deviation of velocities, \sigmavb, as a
function of effective temperature for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of {\it effective
temperature}.
We chose to use effective temperatures and not colors in this analysis as it
is the linear quantity and therefore easier to divide into bins of roughly
equal numbers of stars.

@@

\subsection{Mass-dependent heating}

@@

\subsection{The period gap}
An unsolved mystery within gyrochronology is the origin of the rotation period
gap, first identified by \citet{mcquillan2013}.
This gap can be seen in figure \ref{fig:age_cut} as an under-density of points
between the 0.7-1.0 and 1.0-1.5 Gyr age ranges and roughly follows a line of
constant gyrochronal age of around 1.1 Gyr \citep[according to
the gyrochronology relation of][]{angus2019}.
Several explanations for its origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018}, a rapid change in magnetic field structure
\citep{reinhold2019}, and erroneous rotation period measurements that are
incorrect by a factor of two \citep{koen2018}.
The latter explanation can be ruled out because stars below the gap have
smaller velocity dispersions than the stars above the gap, indicating that
they are kinematically younger \citep{mcquillan2013, davenport2018}, as
evident in figure \ref{fig:age_cut}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
dispersion as the hottest stars in the age range above which indicates that
the period-\teff\ relations are {\it flat} at these rotation periods.
It may be a coincidence that the gyrochronology relations seem to only flatten
off {\it above} the period gap, or it may be that the period gap is located at
magnetic transition period.
We lack a sufficient quantity of data to investigate in detail but new field
star rotation periods from the \ktwo\ and \tess\ missions may be able to
validate or rule out this hypothesis in the future.

@@

% If we assume the rise in velocity dispersion at cooler effective temperatures
% is caused by an inaccurate period-color relation rather than mass-dependent
% dynamical heating, we can generate a graphical representation of what the
% relations {\it should} look like.

\subsection{The period-\teff\ relations, revealed}

\begin{figure}
  \caption{
Rotation period vs effective temperature for stars in the \mct\ sample,
    colored by the velocity dispersions of stars in restricted period and
    temperature ranges.
}
  \centering
    \includegraphics[width=1\textwidth]{dispersion_period_teff}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion.
Here, velocity dispersion was calculated for groups of stars in
uniformly-spaced bins in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating doesn't affect this sample, and \vb\
at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, it appears that rotation period {\it decreases} with
decreasing effective temperature at old ages.  This is a paradigm shift
for gyrochronology because stellar spin-down rate is thought to be directly
tied to magnetic field strength, and the deeper convection zones of cooler
stars generate stronger magnetic fields which {\it should} lead to more
efficient angular momentum loss.
A similar phenomenon has been observed in the 1.1 Gyr NGC 6811 cluster, where
the rotation periods of mid-K dwarfs are faster than expected; their
rotational evolution appears to have stalled, and the period-\teff\ relation
is flat \citep{curtis2019}.
NGC 6811 straddles the rotation period gap and may be the `missing link' that
connects two epoch of stellar spin-down: an early stage where the
period-\teff\ relation for K dwarfs has a negative slope and a late stage
where it has a positive slope.
In this case the period gap may delineate the transition between these two
regimes and is the point at which stellar magnetic dynamos likely undergo a
dramatic structural shift.

Figure \ref{fig:dispersion_period_teff} also shows that the (kinematically)
oldest stars with measured rotation periods are cooler than 4500 K.
This may be because cooler stars remain active for longer, and their rotation
periods can therefore be measured at older ages.
